### 2017-02-08 manuscript.rmd
figure formatting, typos

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document:
    dev: png
    fig_caption: yes
    fig_width: 6
    fig_height: 4
  github_document:
    fig_caption: yes

@@

# Multiple pairwise comparisons between species strongly mislead functional genomics studies

@@

With the goal of providing an integrated understanding of genome and phenotype evolution, there is now considerable interest in comparing functional genomics data across species. Most such studies have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

![Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.png?raw=true)

@@

	g_expectation_null =
		sim_expectation() %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly(  binwidth=0.2, position="identity" ) +
			xlab( "Phylogenetic independent contrasts (PIC)" ) +
			ylab( "Density" ) +
			xlim( 0, 6 ) + 
			theme_classic() +
			theme( 
				axis.text.x=element_blank(), 
				axis.ticks.x=element_blank(), 
				axis.text.y=element_blank(), 
				axis.ticks.y=element_blank(),
				legend.justification=c( 1, 1 ), 
				legend.position=c( 1, 1 )
			) +
			ggtitle( "(a) Null expectation" )

	g_expectation_oc =
		sim_expectation( scale=2 ) %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly(  binwidth=0.2, position="identity" ) +
			xlab( "Phylogenetic independent contrasts (PIC)" ) +
			ylab( "Density" ) +
			xlim( 0, 6 ) + 
			theme_classic() +
			theme( 
				axis.text.x=element_blank(), 
				axis.ticks.x=element_blank(), 
				axis.text.y=element_blank(), 
				axis.ticks.y=element_blank(),
				legend.position="none"
			) +
			ggtitle( "(b) Ortholog Conjecture (OC) expectation" )

@@

	# Plot the contrasts for duplication events and speciation events
	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme_classic()

@@

```{r ascertainment, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

	g_age =
		nodes_contrast %>%
		ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
			geom_point( alpha=0.2 ) + 
			xlab( "Node age (millions of years)" ) + 
			ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
			ylim( 0, 0.3 ) +
			xlim( 0, 1000 ) +
			ggtitle( "(a)" ) +
			theme_classic()

@@

```{r estimate_tau_model, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}
	# Estimate model parameters and add them to the tree objects
	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
	# This takes a while, so do it once here and reuse the values as needed
	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
```

```{r simulations_null, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

```{r simulations_oc, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure 4). There are several difficulty with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

Their Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 9). A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have hhigher maximum expression and lower  This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

@@

```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="Investigation of possible ascertainment biases. (a) Magnitude of independent contrasts plotted against node age. Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range. (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes. (c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths. Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes. (d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}

@@

	g_density_filtered_age =
		nodes_contrast_filtered_age %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent \ncontrasts (PIC)" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		ggtitle( "(b)" ) +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
		theme_classic()

	# Examine by expected variance (branch length)
	g_var_exp =
		nodes_contrast %>%
		ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
		geom_point( alpha=0.2 ) + 
		xlab( "Expected variance, \n depends mostly on branch length" ) + 
		ylab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
		ylim( 0, 0.3 ) +
		xlim( 0, 2000 ) +
		ggtitle( "(c)" ) +
		theme_classic()

@@

	g_density_filtered_var =
		nodes_contrast_filtered_var %>% 
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) + 
		ggtitle( "(d)" ) +
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
		xlim( 0, 0.075 ) + 
		theme_classic() +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
		xlab( "Phylogenetic independent contrasts (PIC)\n of differential expression" ) +
		ylab( "Density" )

@@

```{r calibration_sensitivity, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

@@

```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", fig.cap="There is a negaative relationship between Tau and the maximum observed expression for each gene."}

@@

	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
		geom_point(alpha=0.05) + 
		geom_smooth(method="lm") +
		xlab("Maximum Expression") +
		theme_classic()

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
output:
  pdf_document:
    fig_caption: yes
    fig_width: 6
    fig_height: 4

@@

# Multiple pairwise comparisons between species can strongly mislead functional genomics studies

@@

There is now considerable interest in comparing functional genomics data across species. One goal of this work is to provide an integrated understanding of genome and phenotype evolution. Most studies of this type have relied on multiple pairwise comparisons between species, an approach that does not incorporate information about the evolutionary relationships among species. These pairwise approaches have the potential to mislead conclusions about the evolution of genome function, and are a missed opportunity to learn about biology that can only be understood in an explicit phylogenetic context. Here we examine two recently published studies that compare expression across species with pairwise methods.  We find problems with both that call their conclusions into question. One study interpreted higher expression correlation between orthologs than paralogs as evidence for the ortholog conjecture, *i.e.* the hypothesis that gene function evolution is more rapid after duplication than speciation. Instead, we find that this pattern is due to the structure of the gene phylogenies, not different rates of expression evolution, and is the expected pattern when rates are the same following duplication and speciation. The second study interpreted pairwise comparisons of embryonic gene expression across distantly related animals as evidence for a distinct evolutionary process that gave rise to animal phyla. Instead, we find that the pattern they identified is due to unique feature of a single species that impact multiple pairwise comparisons that include that species. In each study, distinct patterns of pairwise similarity among species were interpreted as evidence of particular evolutionary processes, but instead reflect the structure of phylogenies. These reanalyses concretely demonstrate the inadequacy of pairwise comparisons for analyzing functional genomic data across species, and indicate that it will be critical to adopt phylogenetic comparative methods in future functional genomics work. Fortunately, phylogenetic comparative biology is also a rapidly advancing field with many methods that can be directly applied to functional genomic data.

@@

![Pairwise and phylogenetic comparative approaches. (a) Many comparative functional genomic studies rely on pairwise comparisons, where traits of each gene are compared to traits of other genes (such as orthologs and paralogs). This leads to many more comparisons than unique observations, making each comparison dependent on others. (b) Comparative phylogenetic methods, including phylogenetic independent contrasts [@Felsenstein:1985ua], make a smaller number of independent comparisons, where each contrast measures independent changes along different branches. Phylogenetic approaches are rarely used for functional genomics studies.](Figure_overview.pdf)

@@

	g_expectation_null =
		sim_expectation() %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly(  binwidth=0.2, position="identity" ) +
			xlab( "Phylogenetic independent contrasts" ) +
			ylab( "Density" ) +
			xlim( 0, 6 ) + 
			theme_classic() +
			theme( 
				axis.text.x=element_blank(), 
				axis.ticks.x=element_blank(), 
				axis.text.y=element_blank(), 
				axis.ticks.y=element_blank(),
				legend.justification=c( 1, 1 ), 
				legend.position=c( 1, 1 )
			) +
			ggtitle( "(a) Null expectation" )

	g_expectation_oc =
		sim_expectation( scale=2 ) %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
			geom_freqpoly(  binwidth=0.2, position="identity" ) +
			xlab( "Phylogenetic independent contrasts" ) +
			ylab( "Density" ) +
			xlim( 0, 6 ) + 
			theme_classic() +
			theme( 
				axis.text.x=element_blank(), 
				axis.ticks.x=element_blank(), 
				axis.text.y=element_blank(), 
				axis.ticks.y=element_blank(),
				legend.position="none"
			) +
			ggtitle( "(b) Ortholog Conjecture \nexpectation" )

@@

	# Plot the contrasts for duplication events and speciation events
	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent contrasts" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
		theme_classic()

@@

```{r ascertainment, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

	g_age =
		nodes_contrast %>%
		ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
			geom_point( alpha=0.2 ) + 
			xlab( "Node age (millions of years)" ) + 
			ylab( "Phylogenetic independent \ncontrast" ) +
			ylim( 0, 0.3 ) +
			xlim( 0, 1000 ) +
			ggtitle( "(a)" ) +
			theme_classic()	+
			theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

@@

```{r estimate_tau_model, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
	# Estimate model parameters and add them to the tree objects
	# Subcalls to fitContinuous() are parallel, so don't use mclapply here
	# This takes a while, so do it once here and reuse the values as needed
	gene_trees_calibrated = lapply( gene_trees_calibrated, add_model_parameters )
```

```{r simulations_null, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

```{r simulations_oc, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

```{r simulations_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Analyses of simulated data reveal that pairwise methods produce similar results under both the null model and the ortholog conjecture, while the phylogenetic method distinguishes between the two. (a) Pairwise analysis of dataset simulated under the null model that there is no difference in the rate of expresison evolution following duplication and speciation. (b) Phylogenetic analysis of data simulated under the null model. (c) Pairwise analysis of dataset simulated under the ortholog conjecture, in which there is a higher rate of expresison evolution following duplication than speciation. This plot is very similar to the corresponding plot for the null model. (d) Phylogenetic analysis of data simulated under the ortholog conjecture. This plot is distinct from that for the null model in that phylogenetic independent contrasts for duplication events are significantly larger than after speciation events."}

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2, they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher tau correlation than paralogs. We confirmed that we can replicate this result (Figure 4). There are several difficulties with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

Their Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (Supplementary Figure 9). A simpler explanation for this plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

@@

```{r ascertainment_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.cap="Investigation of possible ascertainment biases. (a) Magnitude of independent contrasts plotted against node age. Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range. (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes. (c) Magnitude of independent contrasts plotted against expected variance, which is largely determined by branch lengths. Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes. (d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes."}

@@

	g_density_filtered_age =
		nodes_contrast_filtered_age %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) +
		xlab( "Phylogenetic independent contrast" ) +
		ylab( "Density" ) +
		xlim( 0, 0.075 ) + 
		ggtitle( "(b)" ) +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
		theme_classic() +
		theme(legend.position="none")

	# Examine by expected variance (branch length)
	g_var_exp =
		nodes_contrast %>%
		ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
		geom_point( alpha=0.2 ) + 
		xlab( "Expected variance, \n depends mostly on branch length" ) + 
		ylab( "Phylogenetic independent \ncontrast" ) +
		ylim( 0, 0.3 ) +
		xlim( 0, 2000 ) +
		ggtitle( "(c)" ) +
		theme_classic()	+
		theme(legend.position="none")

@@

	g_density_filtered_var =
		nodes_contrast_filtered_var %>% 
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
		geom_freqpoly( binwidth=.001, position="identity" ) + 
		ggtitle( "(d)" ) +
		theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
		xlim( 0, 0.075 ) + 
		theme_classic() +
		annotate( "text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
		xlab( "Phylogenetic independent contrast" ) +
		ylab( "Density" ) +
		theme(legend.position="none")

@@

```{r calibration_sensitivity, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

@@

```{r Tau_v_max, cache=TRUE, echo=FALSE, warning=FALSE, dev="png", dpi=300, message=FALSE, fig.cap="There is a negative relationship between Tau and the maximum observed expression for each gene."}

@@

	ggplot( nodes_tips, aes( x=max_expression, y=Tau  ) ) + 
		geom_point( size=0.5, alpha = 0.05 ) + 
		geom_smooth( method="lm" ) +
		xlab( "Maximum Expression" ) +
		theme_classic()
